---
title: "Severe weather types and impact on economy and health"
author: "Ronald Hensbergen"
date: "14 november 2016"
output: html_document
---     

This project is about investigation of storm and other severe weather types that occured between 1950 and 2011. Plan is to identify which types are most devastating economically and to public health.

The data source is the U.S. National Oceanic and Atmospheric Administration's (NOAA) [storm database] [1]. The 1950-2011 data is archived at this [link] [2]

Some documentation is made available at below links:
National Weather Service Storm Data Documentation [Construction/definition of variables] [3]
National Climatic Data Center Storm Events [FAQ] [4]

```{r init,echo=FALSE}
library(ggplot2)

if (!file.exists("./data")) {
        dir.create("./data")
}
fileurl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "./data/stormdata.bz2"
if (!exists("stormdata")) {
        download.file(fileurl, destfile)
        stormdata <- read.csv(destfile)
}
```

```{r first_question}
stormdata$TOTFATINJ <- stormdata$FATALITIES + stormdata$INJURIES
stormdata.pop.mean <-
aggregate(TOTFATINJ ~ EVTYPE, data = stormdata, mean)
stormdata.pop.mean <-
head(stormdata.pop.mean[order(-stormdata.pop.mean$TOTFATINJ),], 4)
stormdata.pop.mean$EVTYPE <-
factor(stormdata.pop.mean$EVTYPE, levels = stormdata.pop.mean$EVTYPE[order(-stormdata.pop.mean$TOTFATINJ)])
g <- ggplot(stormdata.pop.mean, aes(EVTYPE, TOTFATINJ))
g <- g + geom_col()
g <-
g + ggtitle("Four most harmful storm types for personal health")
g <-
g + labs(x = "Storm type", y = "Average of Fatalities and Injuries")
g <- g + theme(title = element_text(size = 14))
g <- g + theme(plot.title = element_text(hjust = 0.5))
g
```

```{r second_question}
stormdata$TOTDMG <- stormdata$PROPDMG + stormdata$CROPDMG
stormdata.econ.mean <-
aggregate(TOTDMG ~ EVTYPE, data = stormdata, mean)
stormdata.econ.mean <-
head(stormdata.econ.mean[order(-stormdata.econ.mean$TOTDMG),], 4)
stormdata.econ.mean$EVTYPE <-
factor(stormdata.econ.mean$EVTYPE, levels = stormdata.econ.mean$EVTYPE[order(-stormdata.econ.mean$TOTDMG)])

g <- ggplot(stormdata.econ.mean, aes(EVTYPE, TOTDMG))
g <- g + geom_col()
g <-
g + ggtitle("Four most harmful storm types financially (property and crops)")
g <- g + labs(x = "Storm type", y = "Total Damage (in $)")
g <- g + theme(title = element_text(size = 14))
g <- g + theme(plot.title = element_text(hjust = 0.5))
g
```

```{r session_info, echo=FALSE}
library(devtools)
devtools::session_info()
```

```{r finishing, echo=FALSE}
```
[1]: https://www.ncdc.noaa.gov/stormevents/
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
[4]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
